---
title: "Predator Problem"
author: "lyzs90"
date: "1 January 2016"
output: html_document
---

## Introduction

In the past three months, there has been 19 [documented] incidents of cat abuse in Yishun. Most of the cats were found dead. Even after the arrest of a suspect on 27 Dec, there was yet another incident. So I've decided to do some rudimentary geographic profiling. My goal was to try and see if more than one culprit has been preying on the cats all along or if there has been an emergence of a copycat killer. Hopefully, I want to also attempt to identify the probable area of residence of the culprit(s).    


## Data Preparation

Referencing [this](http://news.asiaone.com/sites/default/files/cat5.jpg) map done up by the Straits Times, I sourced for the geographic coordinates of all the incidents on Google Maps and keyed the data onto a spreadsheet. Basically, the features I extracted are longitude and latitude, both in decimal degrees. I then normalized the features.

```{r, echo=F, message=F, warning=F}
library(dplyr)
cat <- read.csv("cat.csv", header=T)
cat$norm_lon <- scale(cat$longitude)
cat$norm_lat <- scale(cat$latitude)
df <- cat %>%
        select(norm_lon, norm_lat)
head(df)
```

## How many culprits are there?

I would suppose that the incidents usually take place at night and public transport would be unavailable. So if the offender walks the distance he can cover from his home is going to be limited. Unless he cycles, then it's quite possible to cover the entire area if he stays somewhere central. However, we know that the first suspect lives at Block 115b and it's highly unlikely that he will travel to Block 355 for instance. Hence, I would think that there are multiple culprits even prior to Dec 27. So for a start, I'm going to use the off-the-shelf k-means algorithm to look for clusters in our data.

First, I determine optimal number of clusters to search for using the NbClust package. This passes the features through 30 indices each suggesting a cluster and then by taking a majority vote, the optimal number of clusters is recommended. 

```{r, echo=F, message=F, warning=F, results='hide', fig.keep='last'}
library(NbClust)
set.seed(12345)
nc <- NbClust(df, min.nc=2, max.nc=5, method="kmeans")
par(mfrow=c(1,1)) # reset plotting
barplot(table(nc$Best.n[1,]),
        xlab="Numer of Clusters", ylab="Number of Indices",
        main="Number of Clusters Chosen by 30 Indices")
```

Given a range of 2-5 clusters to work with, NbClust recommends that we use 3 clusters. Does this mean that there may potentially be 3 culprits? 

## Where are the potential bases?

The resultant clusters are shown below. 

```{r, echo=F, message=F, warning=F}
set.seed(12345)
fit.km <- kmeans(df, 3, nstart=25) # do 25 different starts and choose the best
cat$labels <- as.factor(fit.km$cluster)

## obtain unscaled clust means
clust.means <- aggregate(cat[2:3], by=list(cluster=fit.km$cluster), mean)
colnames(clust.means)[1] <- "labels"
clust.means$labels <- as.factor(clust.means$labels)

## plot on map
library(ggmap)
library(ggplot2)
# draw boundary box
lat <- c(1.405, 1.45)                
lon <- c(103.835, 103.84)   
map.loc <- get_map(location = c(lon = mean(lon), lat = mean(lat)),
                maptype = "terrain",
                source = "google",
                color = "bw",
                zoom=15)
ggmap(map.loc, extent = 'device')+
    geom_point(data = cat, aes(x = longitude, y = latitude, color = labels), size=4)+
    geom_point(data = clust.means, aes(x = longitude, y = latitude, color=labels),
               size=8,
               shape=4)+
    scale_colour_brewer("Cluster", palette = "Dark2")
```

The clusters are even sized (6-7 incidents apiece). The purple cluster center (denoted by the X) is close to the first suspects home. However, I also noted that the algorithm incorrectly assigned the latest incident on 30 Dec to the purple cluster. Perhaps including the incident date in our feature set could prevent this erroneous assignment?

## Next Steps

As you can see, the analysis requires more tweaking. For those who are interested, the csv can be found on my GitHub [repo](https://github.com/lyzs90/PredatorProblem). Do let me know if you have any feedback or suggestions. Meanwhile to all Yishunites, please help keep a lookout for our feline friends.